\section{About the book}
This book is a tutorial for the Meruem programming language, written by the people who developed the current version of Meruem. Our goal is to teach you the introductory concepts of functional programming and (to some extent) metaprogramming using the Meruem language. 

\section{Who should read this book}
If you are someone looking for the next popular object-oriented programming language to master and doesn't feel like learning new and more mathematical ways of solving problems for now, then this tutorial is not for you. There are many options for new such languages out there, but Meruem is not (yet) one of them.

That said, if you are someone willing to spend a lot of time mastering not just a new programming language but also different programming paradigms, hoping that you will be able to apply all the knowledge you can gain from this book with any other programming languages you already know, then this book is for you.

\section{Programming background required}
This book is written primarily for imperative and/or object-oriented programmers who want to learn functional programming and metaprogramming, or people who don't know programming at all. If you are already familiar with functional programming and metaprogramming, then most of the contents here will not be new to you, but this book can still serve as a review material.

\section{How to read this book}
Most of the chapters in this book are not self-containing, so we recommend you read them in the proper order starting from chapter 1, especially if you are new to Lisp-like languages like Meruem. 